---
title: "Quarto Starter Kit for IR Professionals"
author: "Colorado State University Office of IRPE"
date: today
format:
  pdf:
    toc: true
    toc_depth: 3
    number-sections: true
    include-before-body: before-body.md
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
  docx:
    toc: true
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Introduction to Quarto

This document serves as both a demonstration of Quarto's capabilities and a reference guide for Institutional Research professionals. You can use the rendered document as a guide, and also examine the source code to see how each feature is implemented.

## What is Quarto?

Quarto is an open-source scientific and technical publishing system built on Pandoc. It allows you to create dynamic documents that can include:

- Formatted text with Markdown
- Executable code (R, Python, Julia)
- Interactive visualizations
- Mathematical equations
- Citations and references

Best of all, from a single source file (like this one), you can produce documents in multiple formats: HTML, PDF, Word, PowerPoint, and more.

```{r}
#| label: setup
#| include: false

# Load essential libraries
library(tidyverse)
library(knitr)
library(flextable)

# Set default options
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

# Getting Started with Quarto

## Installation Instructions

To use Quarto, you'll need to install:

1. R from [CRAN](https://cran.r-project.org/)
2. RStudio from [Posit](https://posit.co/download/rstudio-desktop/)
3. Quarto from [Quarto.org](https://quarto.org/docs/get-started/)

Newer versions of RStudio (2022.07 or later) include Quarto by default.

## Essential R Packages

These packages are commonly used in IR reporting workflows:

```{r}
#| label: package-list
#| eval: false

# Basic packages for data handling and visualization
install.packages(c(
  "tidyverse",    # Data manipulation and visualization
  "knitr",        # Dynamic report generation
  "rmarkdown",    # Markdown processing
  "flextable"     # Word-friendly tables
))

# You only need to install packages once on your computer
```

Remember: You only need to install packages once, but you need to load them with `library()` in each document.

```{r}
#| label: library-example
#| eval: false

library(tidyverse)
library(knitr)
library(rmarkdown)
library(flextable)
```

# Basic Document Features

## Text Formatting with Markdown

Quarto uses Markdown for text formatting:

- **Bold text** with `**bold**`
- *Italic text* with `*italic*`
- Lists with `-` or `1.`
- [Links](https://quarto.org) with `[text](url)`
- Headers with `#`, `##`, etc.

## Code Chunks and Output

This is a code chunk that runs R code and displays its output:

```{r}
#| label: simple-calculation

# Calculating summary statistics for a sample dataset
numbers <- c(15, 23, 42, 56, 89, 12)

# Summary statistics
mean_value <- mean(numbers)
median_value <- median(numbers)
std_dev <- sd(numbers)

# Create a table of results
data.frame(
  Statistic = c("Mean", "Median", "Standard Deviation"),
  Value = c(mean_value, median_value, std_dev)
)
```

## Basic Charts

Creating visualizations is straightforward. Here's a simple bar chart:

```{r}
#| label: simple-chart
#| fig-cap: "Sample enrollment data by college"

# Create sample data
enrollment <- data.frame(
  College = c("Arts & Sciences", "Business", "Engineering", "Education", "Health Sciences"),
  Students = c(5234, 3211, 2543, 1432, 2211)
)

# Create a bar chart
ggplot(enrollment, aes(x = reorder(College, -Students), y = Students)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Enrollment by College",
    x = "College",
    y = "Number of Students"
  ) +
  # Add data labels on top of bars
  geom_text(aes(label = Students), vjust = -0.5) +
  # Rotate x-axis labels for better readability
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Tables

You can create nicely formatted tables for different output formats:

```{r}
#| label: simple-table

# Create sample retention data
retention <- data.frame(
  College = c("Arts & Sciences", "Business", "Engineering", "Education", "Health Sciences"),
  "First Spring" = c(0.82, 0.85, 0.88, 0.79, 0.84),
  "Second Fall" = c(0.76, 0.78, 0.82, 0.72, 0.77),
  "Third Fall" = c(0.71, 0.74, 0.79, 0.68, 0.72)
)

# For HTML and PDF output, kable works well
kable(retention, digits = 2, caption = "Retention Rates by College")
```

For Word output, the flextable package provides better control:

```{r}
#| label: flextable-example

flextable(retention) %>%
  colformat_double(j = 2:4, digits = 2) %>%
  autofit() %>%
  add_footer_lines("Source: Office of Institutional Research") %>%
  theme_vanilla()
```

# Creating Parameterized Reports

## Setting Up Parameters

Parameters allow you to create one template that generates many different outputs. Parameters are defined in the YAML header:

```yaml
---
title: "College of {$college} Fact Sheet"
params:
  college: "Arts & Sciences"
  academic_year: 2024
---
```

## Using Parameters in Your Document

You can refer to parameters in both text and code:

```{r}
#| label: param-demo
#| echo: true

# Define sample parameters (in a real parameterized document, these would come from the YAML header)
params <- list(
  college = "Engineering",
  academic_year = 2024
)

# Display parameter value in text using inline R code
cat(paste("This report shows data for the", params$college, "college in the", params$academic_year, "academic year."))

# Use parameters in data filtering
colleges <- data.frame(
  College = c("Arts & Sciences", "Business", "Engineering", "Education", "Health Sciences"),
  Students = c(5234, 3211, 2543, 1432, 2211),
  Year = rep(2024, 5)
)

# Filter data based on parameter
filtered_data <- colleges %>%
  filter(College == params$college, Year == params$academic_year)

# Show the filtered data
filtered_data
```

## Rendering with Parameters

In RStudio, you can render with parameters by:

1. Clicking the ▼ next to the "Render" button
2. Choosing "Render with Parameters..."
3. Entering your parameter values
4. Clicking "Render"

For batch rendering, you can use an R script:

```{r}
#| eval: false

library(quarto)

# List colleges to generate reports for
colleges <- c("Arts & Sciences", "Business", "Engineering")

# Create a report for each college
for(college in colleges) {
  quarto::quarto_render(
    "college_report.qmd",  # Your template file
    output_file = paste0(college, "_Report.pdf"),
    execute_params = list(
      college = college,
      academic_year = 2024
    )
  )
}
```

# Example: Enrollment Report

Let's create a more complete example combining several features:

```{r}
#| label: dashboard-data

# Create sample enrollment trend data
set.seed(42)
years <- 2020:2024
colleges <- c("Arts & Sciences", "Business", "Engineering", "Education", "Health Sciences")

# Expand grid to create all combinations
trend_data <- expand.grid(
  Year = years,
  College = colleges
)

# Generate reasonably realistic enrollment numbers with some trends
base_enrollments <- c(6000, 3500, 2800, 1800, 2700)
names(base_enrollments) <- colleges

# Create enrollment with modest growth for each college
trend_data$Enrollment <- sapply(1:nrow(trend_data), function(i) {
  college <- as.character(trend_data$College[i])
  year_index <- which(years == trend_data$Year[i])
  
  # Base value plus growth and random noise
  base <- base_enrollments[college]
  growth <- base * 0.02 * (year_index - 1)  # 2% growth per year
  noise <- rnorm(1, 0, base * 0.01)  # Small random component
  
  # Special case: make Education decline slightly
  if (college == "Education") {
    growth <- -growth * 0.5
  }
  
  return(round(base + growth + noise))
})
```

## Enrollment Trends Visualization

```{r}
#| label: dashboard-viz
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Five-Year Enrollment Trends by College"

# Create the line chart
ggplot(trend_data, aes(x = Year, y = Enrollment, color = College, group = College)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = years) +
  theme_minimal() +
  labs(
    title = "Enrollment Trends by College (2020-2024)",
    y = "Total Enrollment",
    x = "Academic Year"
  ) +
  # Add data labels for 2024
  geom_text(
    data = trend_data %>% filter(Year == 2024),
    aes(label = Enrollment),
    nudge_x = 0.2,
    size = 3.5
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Enrollment Summary Table with Calculations

```{r}
#| label: dashboard-table

# Calculate summary statistics
summary_stats <- trend_data %>%
  group_by(College) %>%
  summarize(
    `2024 Enrollment` = Enrollment[Year == 2024],
    `2020 Enrollment` = Enrollment[Year == 2020],
    `Change` = `2024 Enrollment` - `2020 Enrollment`,
    `Percent Change` = ((`2024 Enrollment` - `2020 Enrollment`) / `2020 Enrollment`) *100
  ) %>%
  arrange(desc(`2024 Enrollment`))

# Create a nicely formatted table
flextable(summary_stats) %>%
  colformat_double(j = 5, digits = 1) %>%
  bg(j = 4, bg = function(x) {
    ifelse(x > 0, "#e6f3e6", ifelse(x < 0, "#f7e6e6", "white"))
  }) %>%
  bg(j = 5, bg = function(x) {
    ifelse(x > 0, "#e6f3e6", ifelse(x < 0, "#f7e6e6", "white"))
  }) %>%
  bold(j = 1) %>%
  add_footer_lines("Source: Office of Institutional Research") %>%
  autofit()
```

# Automation Techniques

## File Organization

For efficient reporting workflows, organize your files in a clear structure:

```
InstitutionalResearch/
  ├── Templates/
  │   ├── college_factsheet.qmd
  │   ├── enrollment_report.qmd
  │   └── generate_reports.R
  │
  └── Reports/
      ├── 2024_Spring/
      │   ├── enrollment_summary.pdf
      │   └── college_factsheets/
      └── 2024_Fall/
```

## Automation Script

Here's a simple script for automatically generating multiple reports:

```{r}
#| eval: false

# Simple script to generate multiple reports
library(quarto)

# List the colleges you want to create reports for
colleges <- c("Arts & Sciences", "Business", "Engineering")
academic_year <- 2024

# Create a folder for all the reports
output_folder <- paste0("Factbook_", academic_year)
dir.create(output_folder, showWarnings = FALSE)

# First, create the main enrollment report
print("Creating main enrollment report...")
quarto::quarto_render(
  "enrollment_report.qmd",
  output_file = paste0(output_folder, "/main_enrollment_report.pdf")
)

# Then create a report for each college
for (college in colleges) {
  # Show progress in the console
  print(paste("Creating fact sheet for", college))
  
  # Make a simple filename
  clean_name <- gsub(" |&", "_", tolower(college))
  filename <- paste0(output_folder, "/", clean_name, "_factsheet.pdf")
  
  # Create the report with parameters
  quarto::quarto_render(
    "college_fact_sheet.qmd",
    output_file = filename,
    execute_params = list(
      college = college,
      academic_year = academic_year
    )
  )
}

print("✓ All reports created successfully!")
```

## Scheduling (Windows)

For Windows users, you can schedule your automation script using Task Scheduler:

1. Create a `.bat` file with the following content:
```
"C:\Program Files\R\R-4.2.1\bin\Rscript.exe" "C:\path\to\your\generate_reports.R"
```

2. Schedule this using Windows Task Scheduler to run monthly or as needed

# Troubleshooting Tips

## Common Issues and Solutions

Here are solutions to common problems:

### PDF Output Problems

If you encounter errors with PDF rendering:

```{r}
#| eval: false

# Install TinyTeX (a lightweight LaTeX system)
install.packages("tinytex")
tinytex::install_tinytex()
```

### Table Formatting in Word

For better-looking tables in Word documents, use flextable instead of kable:

```{r}
#| eval: false

library(flextable)
flextable(your_data) %>% autofit()
```

### Data Loading Errors

Always check file paths and preview your data:

```{r}
#| eval: false

# Print working directory to check where R is looking for files
print(getwd())

# Preview first few rows of CSV to check structure
head(read.csv("your_file.csv"))
```

# Resources for Learning More

## Beginner-Friendly Resources

- [Posit's "Getting Started with Quarto"](https://quarto.org/docs/get-started/)
- [R for the Rest of Us](https://rfortherestofus.com/)
- [Introduction to R for Excel Users](https://rstudio-conf-2022.github.io/excel-to-r/)

## Video Tutorials

- [RStudio YouTube Channel](https://www.youtube.com/c/RStudioPBC/videos)
- [How to Create Your First Quarto Document](https://www.youtube.com/watch?v=_f3latmOhew)

## Getting Help

- [Posit Community Forum](https://community.rstudio.com/)
- [Stack Overflow with 'quarto' tag](https://stackoverflow.com/questions/tagged/quarto)

# Thank You!

- Nicole Ross, Associated Director, nmvross@colostate.edu
- Lee Tyson Senior Research Analyst, lktyson@colostate.edu
